\textbf{\sc Towards Deliberative Control of Autonomous Robots}

### 2012-02-07 trunk/arch.tex
updates

\subsection{The T-REX architecture}
\label{sec:arch:trex}

@@

The T-REX architecture was designed with the objective to bring 
planning closer to the low-level control of the vehicle while
maintaining its reactivity. Indeed, in most of the architecture where
task planning is embedded, this task remains at a failry high level of
abstraction as the complexity of this task would otherwise impede the
system reactiveness. As a result the planning problem managed {\em in
  situ} remains fairly detached from low-level control and most of the 
reactive adaptation which in contrast is managed by specialized 
component that relies on different representations for modeling 
the local behavior they control or even do not have any explicit 
formal model of this behavior. The two angles our architecture attck
this issue are the following:
\begin{itemize}
\item Distribute the global planning problem into multiple decision
  loops, called {\em reactors}. Each reactor have its own scope both 
  functionally (one for example focusing on the conversion of high 
  level waypoints into low-level commands) and temporally as 
  each reactor specifies to the agent their planning look-ahead and
  their expected latency before being able to produce plan.
\item Enforce a stronger interaction between the planning process 
  and the integration of the world evolution. The objective being to
  make sure that the plannign task is done while continuously informed 
  on new world state change. By doing so, we expect to allow the planner
  tohave better informed decision despite the fact that its latency is
  larger that the world change rate.
\end{itemize}

For the first point T-REX provides a formal framework described in
details in \cite{Py:2010ti} and is not the focus of this paper. The
second have deep implication on how on should integrate a planner in
this framework in general and was applied to the europa framework
specifically. We'll develop later how this requirement implies to see
in-situ planning as two parallels processes that are tightly linked
through the same plan structure. But before that we need to introduce
at least the high leve lconcepts and ideas behind our architecture as
a whole.

T-REX architecture is structured around the notion of 
{\em state variables} as the unique basis of interactionalong
with a well defined ownership  model for each of these. 
In this framework each reactor is seen as a
black box providing its own {\em internal} state variables for which
its is reponsible to maintain its value at every single instant --
represented by a {\em tick} with a fixed duration. In order to
identify its state a reactor can subscribe to {\em external} state
variables for  which it will received new updates when the owner of
these will produce its {\em internal} state. This {\em
  synchronization} process progate through the reactor dependency 
graph on a bottom up flow at every {\em tick} ensuring that all the
reactors have a consistent view of the world. Conversely each reactor
can request a future change on one of its {\em external} state
varaiable which will be transfromed as an {\em internal} goal to the
owner. This goal is expected to porduce a deliberation phase on the
receiver which can in turn produce a plan impacting its {\em external}
state variables that will propagate down following the same
mechanism.

@@

Take for example the T-REX agent instance we use in our AUV
(Fig. \ref{fig:agent}). In this architecture all the reactors
(symbolized by colored boxes) do not have explicit connection from one
to another but instead rely on a publish/subscribe model on state
variables. For Example the \textsf{Pilot} reactor needs the {\em
  externally} managed state informations from \textsf{Sensor\_Data} 
and \textsf{Nav\_cmd} states variables in order to maintain {\em
  internally} the state variables \textsf{Waypoint} and
\textsf{Nav\_policy}. This relation applies in two way :
\begin{itemize}
\item In order to identify its current {\em internal} state values
  this reactor needs to know the current {\em external} state
  information it relies on. In this case the current \textsf{Waypoint}
  the vehicle is heading to can be extracted for example from the
  current \textsf{Nav\_cmd} executed.
\item The future objectives a reactor has on its future \textsf{Internal}
  state will probably imply sub-objectives to be reported to the
  owner of its {\em external} depending on their current values. Again
  if this reactor wants to go ato a specific \textsf{Waypoint} --
  depending on its current position provided by \textsf{Sensor\_Data}
  it can identify a sequence of \textsf{Nav\_cmd} that should
  eventually help him reach this location.
\end{itemize}

@@

Overal T-REX architecture helps to abstract to the reactor interaction
to only state variables. Each reactor is practically ignoring where
its {\em external} state information is managed as long as this
information is available and properly maintained so they can properly 
manage their own {\em internal} state information which in turn may be
used by other reactors. Such formalism also constrain reactors
interaction to state information exchanges which are two folds:
\begin{itemize}
\item {\em Observation on the current state values}: This information
  propagates up in the reactor graph at the synchronization phase 
  and occur at the beginning of every time steps (called
  ticks). During this phase the agent provides to each reactor 
  updates on their {\em external} state variables for this tick and
  ask them to compute their {\em internal} state variables values for
  this same tick. By doing so we propagate a consistent view of that
  state of the world throughout all the reactors as time advance.
\item {\em Goals request on future state}: This information propagates
  down in the reactor graph. It reflects that, in order to satisfy own
  of its {\em internal} objectives, a reactor may have a plan that
  rely on future values of one of its {\em external} state. When such
  a plan is identified we ensure that this part of the plan is given
  as a goal to the owner of the given {\em external(s)} state variable(s). 
\end{itemize}

@@

The overall execution cycle of the agent is also abstracted as some
continuous planning/deliberation cycle between all the
reactors. Indeed , at the agent interface level, each reactor provided
only two abstract calls that conceptualize their execution:
\begin{itemize}
\item \texttt{synchronize} that takes the last {\em external}
  observations as argument and will return the {\em internal}
  observations for this reactor or a failure report.
\item \texttt{step} that accept new goals requested (if any) on the
  {\em internal} timeline of this reactor and will execute one step of
  deliberation. The returned value indicates if a complete plan has
  been found and in that case will also provide the part of this plan
  applied to {\em external} timelines of this reactor.
\end{itemize}
Both of these two deliberative functions have different scope and
temporal constraints. The \texttt{synchronize} call is considered as
atomic and will focus only on state inference from the current tick by
identifying the current value of the {\em internal} state variables
from the latest updates on the {\em external} variables. This process
will then propagate up in the reactor graph at any single tick to
ensure that all reactors are aware of the current state of the
world. \texttt{step} relates to a single decision step of the planning
process which is focused on the future evolution of the state
variables and it can take multiple steps spanning other multiple
ticks for a reactor to produce a plan that is both complete and
valid. when such plan is identified its {\em external} part is then
dispatched to the corresponding reactors which in turn can start their
deliberation \texttt{step}s in order to satisfy these new
objectives. As a consequence physical execution will occur as these
goals and planning propagate down in the reactor graph until they
eventually become goals for the lowest level reactor which -- instead
of deliberating -- may just transform this goal in a physical command
execution.

The way the agent interleave planning \texttt{step}s and
synchronization is illustrated in Figure \ref{fig:tick-exec}.
Synchronization happens at every single tick and will interrupt
planning allowing this reactor to identify its current state that can
propagate through the reactor graph in order to ensure t aconsitent
view of the state of the world between all the reactors.  
\begin{figure}[!htbp]
  \centering
  \vskip-1pc
  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
  \caption{\small \texttt{T-REX} reactor execution cycle: {\em
      Deliberation} is interrupted by {\em synchronization} at the
    beginning of every {\em tick} allowing integration of state
    information.}
  \label{fig:tick-exec}
  \vskip-0.8pc
\end{figure}
This view of the reactor has two parallel tasks aloow for long
planning deliberation while ensuring that the state evolution
propagates through the agent as time advance. These two tasks 
intertwined together give the two natural flows (bottom-up for state
estimation and top to bottom for future plan) that we expect to see in
a control loop. As they are both conceptualized as inference processes
it is natural to implement a reactor based on generic automated
planning. For example in Fig. \ref{fig:agent}, appart \textsf{Shore
  comm.} and \textsf{Vehicle} acting as interfaces to the system,
every other reactors (colored in blue) are different instance of our
europa reactor differntiated only by their model and europa solver
configurations.

@@

The europa reactor is our implementation inside T-REX of a reactor
that usess automated planning for its execution. This reactor is fully
based on the europa framework and applies the core idea we of T-REX we
described above to implement a reactor for which the sole purpse is to
deliberate as time advance. The choice of europa as the planning
framework was directed by the fact thta this framework provide  a rich
and highly configurable framework to be able to infer around a rich
temporal domain.

This reactor is implemented following the idea that in T-REX, the
whole execution cycle is two deliberation processes. Both of these
processes are implemented as two europa solvers modifying the same
plan database and for which the execution is managed by the T-REX 
framework as time advance during the execution of the system:
\begin{enumerate}
\item The synchronization solver is a specialized europa solver that
  will integrate in the plan new information about the evolution of
  the External state variable and ensure that the reactor propagate
  these in order to both identify its current state and inform
  other reactors of any state change on its internal timelines. This
  solver is summoned at the beginning of every single tick.
\item The deliberation solver is managing the deliberation process of
  the reactor in order to produce a new plan or alter its current plan
  as new goals are given to the reactor or the synchronization solver
  identifying a conflict between the current state of the world and
  the expectations of the previous plan. This process can  span other
  multiple tick and therefore can be interrupted at any single tick in
  order for the synchronization solver to do its task.
\end{enumerate}
While the two processes are separate, they share the same plan
internal to a reactor and therefore at every synchronization cycle the
planning process is informed of the new world state and its impact on
the plan. Conversely, when the planning process eventually finds a
solution, synchronization is informed about generating the planned
state values on the {\em external} state variables. This {\em
  external} plan defines the set of {\em goals} for this state
variable managed by another {\em reactor}.

An important challenge for embedding automayted planning i na situated
agent as always been to go around the asumptions made by most classical
planners. Specifically one assumption restrict the problem to ``offline
planning''  defined in \cite{ghallab04} as follow:
\begin{quotation}
  The planner is not concerned with any change that may occur in
  $\Sigma$ \footnote{In this book $\Sigma$ identifies the world modeled
    by the plan domain} while it is planning; it plans for the given
  initial and goal states regardless of the current dynamics, if any.
\end{quotation}
This assumption, while helpfull to reduce the scope of the planning
problem, becomes problematic if one wants to embed a planner in a
real-time system that evolves in a highly dynamic environment. In a
typical robot this implies that the only point where the planner can
be informed about the current world state is righe before the planning
is started by giving to the planner the intial state and current
objectives. During the planning, it is assumed that the agent can
maintin the world as perceived by the planner as stable. In 
\cite{lemai04, lemai-chenevier2004}, they do attempt to reduce the
impact of planning by aloowing local plan repairs which attempt to
insert repairing actions in the exisiting plan andf can be done fast
enough. Still in ythe eventuality that such repair is not possible
they add to stop their vehicle until the full (re-)planning was
complete. {\em I may need more examples/refs here}

@@

In the europa based reactor we introduce this requirement as a new
type of flaw for the europa framework. This flaw enforces at the
reactor to identify fully its internal state for the current
tick. By using this new flaw we can describe the synchronization
process as this general sequence :
\begin{enumerate}
\item Integrate the external state as provided by the owner of each
  external timeline into the plan database.
\item Propagate this information in the plan database following the
  model $\mathcal{M}$ of this reactor.
\item Resolve the current state value of each internal timelines.
\end{enumerate}

@@

The resolution of internal state flaw can be resolved using one of the
following choices which are evaluated in the given order:
\begin{enumerate}
\item Extend the previous state value to end after this tick (ie
  restrict its end time to $[\tau+1, \infty)$). 
\item Start the next active token in the timeline and
  attempt to start it now (ie restrict its start time to the single
  value $\tau$).
\item Create and insert a new token in this timeline that will start
  at the current tick $\tau$ (attempt this for each possible token
  type for this timeline if necessary). 
\end{enumerate}

All of these choices are evaluated sequentially until a consistent
solution with no more flaw for this tick is identified. In order to do
so we need to make the assumption that the current state value of an
internal timeline do not depend on the future or more accurately
that any choices made during this synchronization will not lead to a
future inconsistency (meaning no possible solution) for a future
synchronization. Such assumption deeply impact the set of possible
domains one the reactor can support while remaining complete. Take for
example the following model rule where {\tt Vehicle} is {\em Internal}
and {\tt Command} is {\em External} to the reactor.
\begin{verbatim}
Vehicle::Dive {
  contains(Command.Descend descend);
  descend.start = start + 2;
}
\end{verbatim}

Such model present the issue that the reactor needs at one point to
take the decision to start the token {\tt Vehicle.Dive} and by doing
so it assumes that in {\em exactly} 2 ticks in the future the {\tt
  Command} timeline will change its state to {\tt Command.Descend}. If
this observation does not occur at the given time this implies that
the reactor estimation was wrong leading him to an inconsistent view
of the world. Conversely the reactor can decide to not switch this
timeline to the {\tt Vehicle.Dive} state and observe 2 ticks in the
future that the {\tt Command.Descend} state value. In this case it
would have missed the opportunity to set its {\tt Vehicle} state to
the correct value which may similarly lead (depending on the rest of
the model) to a similar inconsistency it won't be able to resolve (as
TREX enforces that  the past is monotonic).

@@

One final aspect to keep into account is that in T-REX point of view
synchronization is the only critical task of a reactor. Indeed, such
process is used in order for the agent to identify its
current state and maintain it consistent throughout the agent
life-time. A reactor not being able to find a consistent solution
during synchronization -- meaning that after evaluating all the
possible solutions provided by its model it did not found any plan
that is not inconsistent -- not only expresses that this reactor is
not able anymore to explain the state of the world but may jeopardize
other reactors that relies on its {\em internal} state. For this
reason a failure to do synchronization immediately result on the
agent killing this reactor and notifying any reactor that depends on
its {\em internal} state variables by posting the special observation
\textsf{Failed}. By doing so we allow for graceful degradation where
other reactors can readapt their plan as they receive this
\textsf{Failed} observation.

@@

The planning process inside a reactor is directly based on the way
europa work which was described in previous section. Still some simple
alterations were made in  order to both include deliberation
information one reactor provide to the architecture -- such as its
deliberation latency and its planning look-ahead both expressed in 
ticks -- and the fact that we potentially need to interleave this
deliberation with not the synchronization process but also with 
the deliberation of other reactors\footnote{Current implementation of
  T-REX is running on a single process and it is the responsibility of
  the agent itself to emulate the multi-threading of the reactors
  deliberation and synchronization}. For this reason while the overal
planning of a europa reactor strictly relies on the europa framework
plnanning solver, its slices the execution of Algorithm
\ref{alg:europa:solve} in atomic steps. The interruption is done
around the recursive call (line \ref{li:recurse}) while always
ensuring that the current partial plan mainitained is never poroven
inconsitent at any single step. This can be managed by allowing the
call to backtrack until a DecisionPoInt that is not exhausted (ie
$decision.hasNext()$ is \texttt{true}) is find in the call stack.

@@

The number of flaw between two steps can evolve for the following
reasons that are exogenous to deliberation :
\begin{enumerate}
\item A new goal has been posted on one of the reactor timeline. This
  is usually due to another reactor finding a plan and posting the
  corresponding new objective to this reactor.
\item The outcome of synchronization created new flaws that did not
  need to be resolved during synchronization.
\item A previously requested goal has be recalled by the initial
  requester. This can happen when the reactor did identify that is
  initial plan is no longer valid. For example as synchronization
  proved this plan to be inconsistent.
\end{enumerate}
These 3 events will potentially produce new flaws to be resolved while
any single deliberation step will usually reduce the number of flaws
to be resolved.

The scope of planning for the reactor plays also a role in the
deliberation process as we limit deliberation to the sliding temporal
window $[\tau+\lambda, \tau+\lambda+\pi]$ where:
\begin{itemize}
\item $\tau$ is the current tick
\item $\lambda$ is the specified latency of this reactor. Which is an
  indicator on the expected maximum time this reactor is expected
  before being able to produce a plan.\footnote{Do note that this
    parameter remains indicative and, while it is highly recommended to
    select a sound value, a failure to produce a plan in this delay is
    not considered as a critical failure of the reactor}
\item $\pi$ is the planning look-ahead of the reactor. which indicates
  how many ticks in the future this reactor is looking ahead while planning.
\end{itemize}
This scope allow to filter out any tokens that are either necessarily
ending before $\tau$ or necessarily starting after. Focusing the
[planning problem to only the tokens that are in a reasonable
future. This window is taken into account by the agent which will
notify the reactor of new {\em internal} goals only when they overlap
this window but also at the deliberation solver in order to reduce the
number of tokens to be evaluated and consequently reduce the cost of
deliberation. As time advance flaws that where initially beyond this
window will become active flaw that the reactor can then evaluate
resulting on an apparent continuous planning of the reactor.

When there's no more flaw is present in the current planning scope a
plan  solution is found. This has two effects:
\begin{enumerate}
\item This reactor does not need to deliberate until the next synchronization.
\item The {\em external} part of this plan can then be posted which
  eventually (at the end of the current tick if these goals overlap
  the owner planning window) will be dispatched as goals triggering
  new deliberation on these reactors.
\end{enumerate}

@@

A core aspect of this reactor is that it takes advantage of the
specific sequencing enforced by the T-REX architecture along with the
fact that only one $plan$ structure is maintained in the reactor and
shared between synchronization and planning to allow these two processes
to not only be interleaved but impact each other through the way they
manipulate this plan.

@@

Now, consider that at the next tick $\tau+1$ the reactor managing the
agent location was able to change the location state varaibale to 
the \texttt{Going}. T-REX informs our reactor of this new observation
which is added in the plan structure as a fact -- starting exactly at
$\tau+1$ and lastign for an unlknown duration which is at least 1
tick. As this new observation is integrated in the plan through
synchronization the solver can identify the similarity between this
new observation oand the next token in the existing plan. Which mean
that instead of making a blind exhaustive search of all the possible
implication the synchronization process starts its execution with the
assumption that these new observations are consitent with the current
plan maintained. As a result at the end of synchronization we will
have the plan depicted in Fig. \ref{fig:shop:exec1}. Which just
propagated the information provided by the new observation  in the
exisiting plan (restricting in turn the start time of 
\texttt{Going} and by propagtion of the constraints \texttt{Go} to  be
{\em exactly} $\tau+1$). While focusing uniquely on the execution
frontier ($\tau+1$ in this figure), the europa solver took benefit of
the exisiting plan to evaluate a solution that asumes that the
observation is consistent with what was decided  inside this
reactor.

>>>>

\subsection{The \rx architecture}
\label{sec:arch:trex}

@@

The \rx architecture was designed with the objective to bring planning
closer to the low-level control of the vehicle while maintaining its
reactivity. Indeed, in most of the architecture where task planning is
embedded, this task remains at a failry high level of abstraction as
the complexity of this task would otherwise impede the system
reactiveness. As a result the planning problem managed {\em in situ}
remains fairly detached from low-level control and most of the
reactive adaptation which in contrast is managed by specialized
component that relies on different representations for modeling the
local behavior they control or even do not have any explicit formal
model of this behavior. The two angles our architecture attck this
issue are the following:

\begin{itemize}

\item Distribute the global planning problem into multiple decision
  loops, called {\em reactors}. Each reactor have its own scope both 
  functionally (one for example focusing on the conversion of high 
  level waypoints into low-level commands) and temporally as 
  each reactor specifies to the agent their planning look-ahead and
  their expected latency before being able to produce plan.

\item Enforce a stronger interaction between the planning process 
  and the integration of the world evolution. The objective being to
  make sure that the plannign task is done while continuously informed 
  on new world state change. By doing so, we expect to allow the planner
  tohave better informed decision despite the fact that its latency is
  larger that the world change rate.

\end{itemize}

For the first point \rx provides a formal framework described in
details in \cite{py10} and \cite{rajan12} and is not the focus of this
paper. The second have deep implication on how on should integrate a
planner in this framework in general and was applied to the europa
framework specifically. We'll develop later how this requirement
implies to see in-situ planning as two parallels processes that are
tightly linked through the same plan structure. But before that we
need to introduce at least the high leve lconcepts and ideas behind
our architecture as a whole.

\rx architecture is structured around the notion of {\em state
  variables} as the unique basis of interactionalong with a well
defined ownership model for each of these.  In this framework each
reactor is seen as a black box providing its own {\em internal} state
variables for which its is reponsible to maintain its value at every
single instant -- represented by a {\em tick} with a fixed
duration. In order to identify its state a reactor can subscribe to
{\em external} state variables for which it will received new updates
when the owner of these will produce its {\em internal} state. This
{\em synchronization} process progate through the reactor dependency
graph on a bottom up flow at every {\em tick} ensuring that all the
reactors have a consistent view of the world. Conversely each reactor
can request a future change on one of its {\em external} state
variable which will be transfromed as an {\em internal} goal to the
owner. This goal is expected to porduce a deliberation phase on the
receiver which can in turn produce a plan impacting its {\em external}
state variables that will propagate down following the same mechanism.

@@

Take for example the \rx agent instance we use in our AUV
(Fig. \ref{fig:agent}). In this architecture all the reactors
(symbolized by colored boxes) do not have explicit connection from one
to another but instead rely on a publish/subscribe model on state
variables. For Example the \textsf{Pilot} reactor needs the {\em
  externally} managed state informations from \textsf{Sensor\_Data}
and \textsf{Nav\_cmd} states variables in order to maintain {\em
  internally} the state variables \textsf{Waypoint} and
\textsf{Nav\_policy}. This relation applies in two way :

\begin{itemize}

\item In order to identify its current {\em internal} state values
  this reactor needs to know the current {\em external} state
  information it relies on. In this case the current \textsf{Waypoint}
  the vehicle is heading to can be extracted for example from the
  current \textsf{Nav\_cmd} executed.

\item The future objectives a reactor has on its future \textsf{Internal}
  state will probably imply sub-objectives to be reported to the
  owner of its {\em external} depending on their current values. Again
  if this reactor wants to go ato a specific \textsf{Waypoint} --
  depending on its current position provided by \textsf{Sensor\_Data}
  it can identify a sequence of \textsf{Nav\_cmd} that should
  eventually help him reach this location.

\end{itemize}

@@

Overall the \rx architecture helps to abstract to the reactor
interaction to only state variables. Each reactor is practically
ignoring where its {\em external} state information is managed as long
as this information is available and properly maintained so they can
properly manage their own {\em internal} state information which in
turn may be used by other reactors. Such formalism also constrain
reactors interaction to state information exchanges which are two
folds:

\begin{itemize}

\item {\em Observation on the current state values}: This information
  propagates up in the reactor graph at the synchronization phase 
  and occur at the beginning of every time steps (called
  ticks). During this phase the agent provides to each reactor 
  updates on their {\em external} state variables for this tick and
  ask them to compute their {\em internal} state variables values for
  this same tick. By doing so we propagate a consistent view of that
  state of the world throughout all the reactors as time advance.

\item {\em Goals request on future state}: This information propagates
  down in the reactor graph. It reflects that, in order to satisfy own
  of its {\em internal} objectives, a reactor may have a plan that
  rely on future values of one of its {\em external} state. When such
  a plan is identified we ensure that this part of the plan is given
  as a goal to the owner of the given {\em external(s)} state variable(s).

\end{itemize}

@@

The overall execution cycle of the agent is also abstracted as some
continuous planning/deliberation cycle between all the
reactors. Indeed , at the agent interface level, each reactor provided
only two abstract calls that conceptualize their execution:

\begin{itemize}

\item \texttt{synchronize} that takes the last {\em external}
  observations as argument and will return the {\em internal}
  observations for this reactor or a failure report.

\item \texttt{step} that accept new goals requested (if any) on the
  {\em internal} timeline of this reactor and will execute one step of
  deliberation. The returned value indicates if a complete plan has
  been found and in that case will also provide the part of this plan
  applied to {\em external} timelines of this reactor.

\end{itemize}

Both of these two deliberative functions have different scope and
temporal constraints. The \texttt{synchronize} call is considered as
atomic and will focus only on state inference from the current tick by
identifying the current value of the {\em internal} state variables
from the latest updates on the {\em external} variables. This process
will then propagate up in the reactor graph at any single tick to
ensure that all reactors are aware of the current state of the
world. \texttt{step} relates to a single decision step of the planning
process which is focused on the future evolution of the state
variables and it can take multiple steps spanning other multiple ticks
for a reactor to produce a plan that is both complete and valid. when
such plan is identified its {\em external} part is then dispatched to
the corresponding reactors which in turn can start their deliberation
\texttt{step}s in order to satisfy these new objectives. As a
consequence physical execution will occur as these goals and planning
propagate down in the reactor graph until they eventually become goals
for the lowest level reactor which -- instead of deliberating -- may
just transform this goal in a physical command execution.

The way the agent interleave planning \texttt{step}s and
synchronization is illustrated in Figure \ref{fig:tick-exec}.
Synchronization happens at every single tick and will interrupt
planning allowing this reactor to identify its current state that can
propagate through the reactor graph in order to ensure t aconsitent
view of the state of the world between all the reactors.

\begin{figure}[!htbp]
  \centering
  \vskip-1pc
  \includegraphics[width=0.55\columnwidth]{figs/tick-cycle}
  \caption{\small \texttt{T-REX} reactor execution cycle: {\em
      Deliberation} is interrupted by {\em synchronization} at the
    beginning of every {\em tick} allowing integration of state
    information.}
  \label{fig:tick-exec}
  \vskip-0.8pc
\end{figure}

This view of the reactor has two parallel tasks aloow for long
planning deliberation while ensuring that the state evolution
propagates through the agent as time advance. These two tasks 
intertwined together give the two natural flows (bottom-up for state
estimation and top to bottom for future plan) that we expect to see in
a control loop. As they are both conceptualized as inference processes
it is natural to implement a reactor based on generic automated
planning. For example in Fig. \ref{fig:agent}, appart \textsf{Shore
  comm.} and \textsf{Vehicle} acting as interfaces to the system,
every other reactors (colored in blue) are different instance of our
europa reactor differntiated only by their model and europa solver
configurations.

@@

The \eu reactor is our implementation inside \rx of a reactor that
usess automated planning for its execution. This reactor is fully
based on the europa framework and applies the core idea we of \rx we
described above to implement a reactor for which the sole purpse is to
deliberate as time advance. The choice of europa as the planning
framework was directed by the fact thta this framework provide a rich
and highly configurable framework to be able to infer around a rich
temporal domain.

This reactor is implemented following the idea that in \rx, the
whole execution cycle is two deliberation processes. Both of these
processes are implemented as two europa solvers modifying the same
plan database and for which the execution is managed by the \rx 
framework as time advance during the execution of the system:

\begin{enumerate}

\item The synchronization solver is a specialized europa solver that
  will integrate in the plan new information about the evolution of
  the External state variable and ensure that the reactor propagate
  these in order to both identify its current state and inform
  other reactors of any state change on its internal timelines. This
  solver is summoned at the beginning of every single tick.

\item The deliberation solver is managing the deliberation process of
  the reactor in order to produce a new plan or alter its current plan
  as new goals are given to the reactor or the synchronization solver
  identifying a conflict between the current state of the world and
  the expectations of the previous plan. This process can  span other
  multiple tick and therefore can be interrupted at any single tick in
  order for the synchronization solver to do its task.

\end{enumerate}

While the two processes are separate, they share the same plan
internal to a reactor and therefore at every synchronization cycle the
planning process is informed of the new world state and its impact on
the plan. Conversely, when the planning process eventually finds a
solution, synchronization is informed about generating the planned
state values on the {\em external} state variables. This {\em
  external} plan defines the set of {\em goals} for this state
variable managed by another {\em reactor}.

An important challenge for embedding automayted planning i na situated
agent as always been to go around the asumptions made by most classical
planners. Specifically one assumption restrict the problem to ``offline
planning''  defined in \cite{ghallab04} as follow:

\begin{quotation}
  The planner is not concerned with any change that may occur in
  $\Sigma$ \footnote{In this book $\Sigma$ identifies the world modeled
    by the plan domain} while it is planning; it plans for the given
  initial and goal states regardless of the current dynamics, if any.
\end{quotation}

This assumption, while helpfull to reduce the scope of the planning
problem, becomes problematic if one wants to embed a planner in a
real-time system that evolves in a highly dynamic environment. In a
typical robot this implies that the only point where the planner can
be informed about the current world state is righe before the planning
is started by giving to the planner the intial state and current
objectives. During the planning, it is assumed that the agent can
maintin the world as perceived by the planner as stable. In
\cite{lemai04, lemai-chenevier2004}, they do attempt to reduce the
impact of planning by aloowing local plan repairs which attempt to
insert repairing actions in the exisiting plan andf can be done fast
enough. Still in ythe eventuality that such repair is not possible
they add to stop their vehicle until the full (re-)planning was
complete. {\em I may need more examples/refs here}

@@

In the europa based reactor we introduce this requirement as a new
type of flaw for the europa framework. This flaw enforces at the
reactor to identify fully its internal state for the current tick. By
using this new flaw we can describe the synchronization process as
this general sequence :

\begin{enumerate}

\item Integrate the external state as provided by the owner of each
  external timeline into the plan database.

\item Propagate this information in the plan database following the
  model $\mathcal{M}$ of this reactor.

\item Resolve the current state value of each internal timelines.

\end{enumerate}

@@

The resolution of internal state flaw can be resolved using one of the
following choices which are evaluated in the given order:

\begin{enumerate}

\item Extend the previous state value to end after this tick (ie
  restrict its end time to $[\tau+1, \infty)$).

\item Start the next active token in the timeline and
  attempt to start it now (ie restrict its start time to the single
  value $\tau$).

\item Create and insert a new token in this timeline that will start
  at the current tick $\tau$ (attempt this for each possible token
  type for this timeline if necessary).

\end{enumerate}

All of these choices are evaluated sequentially until a consistent
solution with no more flaw for this tick is identified. In order to do
so we need to make the assumption that the current state value of an
internal timeline do not depend on the future or more accurately that
any choices made during this synchronization will not lead to a future
inconsistency (meaning no possible solution) for a future
synchronization. Such assumption deeply impact the set of possible
domains one the reactor can support while remaining complete. Take for
example the following model rule where {\tt Vehicle} is {\em Internal}
and {\tt Command} is {\em External} to the reactor.

\begin{verbatim}
Vehicle::Dive {
  contains(Command.Descend descend);
  descend.start = start + 2;
}
\end{verbatim}

Such model presents the issue that the reactor needs at one point to
take the decision to start the token {\tt Vehicle.Dive} and by doing
so it assumes that in {\em exactly} 2 ticks in the future the {\tt
  Command} timeline will change its state to {\tt Command.Descend}. If
this observation does not occur at the given time this implies that
the reactor estimation was wrong leading him to an inconsistent view
of the world. Conversely the reactor can decide to not switch this
timeline to the {\tt Vehicle.Dive} state and observe 2 ticks in the
future that the {\tt Command.Descend} state value. In this case it
would have missed the opportunity to set its {\tt Vehicle} state to
the correct value which may similarly lead (depending on the rest of
the model) to a similar inconsistency it won't be able to resolve (as
TREX enforces that the past is monotonic).

@@

One final aspect to keep into account is that in \rx point of view
synchronization is the only critical task of a reactor. Indeed, such
process is used in order for the agent to identify its current state
and maintain it consistent throughout the agent life-time. A reactor
not being able to find a consistent solution during synchronization --
meaning that after evaluating all the possible solutions provided by
its model it did not found any plan that is not inconsistent -- not
only expresses that this reactor is not able anymore to explain the
state of the world but may jeopardize other reactors that relies on
its {\em internal} state. For this reason a failure to do
synchronization immediately result on the agent killing this reactor
and notifying any reactor that depends on its {\em internal} state
variables by posting the special observation \textsf{Failed}. By doing
so we allow for graceful degradation where other reactors can readapt
their plan as they receive this \textsf{Failed} observation.

@@

The planning process inside a reactor is directly based on the way
europa work which was described in previous section. Still some simple
alterations were made in order to both include deliberation
information one reactor provide to the architecture -- such as its
deliberation latency and its planning look-ahead both expressed in
ticks -- and the fact that we potentially need to interleave this
deliberation with not the synchronization process but also with the
deliberation of other reactors\footnote{Current implementation of \rx
  is running on a single process and it is the responsibility of the
  agent itself to emulate the multi-threading of the reactors
  deliberation and synchronization}. For this reason while the overal
planning of a europa reactor strictly relies on the europa framework
plnanning solver, its slices the execution of Algorithm
\ref{alg:europa:solve} in atomic steps. The interruption is done
around the recursive call (line \ref{li:recurse}) while always
ensuring that the current partial plan mainitained is never proven
inconsistant at any single step. This can be managed by allowing the
call to backtrack until a DecisionPoInt that is not exhausted (ie
$decision.hasNext()$ is \texttt{true}) is find in the call stack.

@@

The number of flaws between two steps can evolve for the following
reasons that are exogenous to deliberation :

\begin{enumerate}

\item A new goal has been posted on one of the reactor timeline. This
  is usually due to another reactor finding a plan and posting the
  corresponding new objective to this reactor.

\item The outcome of synchronization created new flaws that did not
  need to be resolved during synchronization.

\item A previously requested goal has be recalled by the initial
  requester. This can happen when the reactor did identify that is
  initial plan is no longer valid. For example as synchronization
  proved this plan to be inconsistent.

\end{enumerate}

These 3 events will potentially produce new flaws to be resolved while
any single deliberation step will usually reduce the number of flaws
to be resolved.

The scope of planning for the reactor plays also a role in the
deliberation process as we limit deliberation to the sliding temporal
window $[\tau+\lambda, \tau+\lambda+\pi]$ where:

\begin{itemize}

\item $\tau$ is the current tick

\item $\lambda$ is the specified latency of this reactor. Which is an
  indicator on the expected maximum time this reactor is expected
  before being able to produce a plan.\footnote{Do note that this
    parameter remains indicative and, while it is highly recommended
    to select a sound value, a failure to produce a plan in this delay
    is not considered as a critical failure of the reactor}

\item $\pi$ is the planning look-ahead of the reactor. which indicates
  how many ticks in the future this reactor is looking ahead while planning.

\end{itemize}

This scope allow to filter out any tokens that are either necessarily
ending before $\tau$ or necessarily starting after. Focusing the
[planning problem to only the tokens that are in a reasonable
future. This window is taken into account by the agent which will
notify the reactor of new {\em internal} goals only when they overlap
this window but also at the deliberation solver in order to reduce the
number of tokens to be evaluated and consequently reduce the cost of
deliberation. As time advance flaws that where initially beyond this
window will become active flaw that the reactor can then evaluate
resulting on an apparent continuous planning of the reactor.

When there are no more flaw is present in the current planning scope a
plan solution is found. This has two effects:

\begin{enumerate}

\item This reactor does not need to deliberate until the next synchronization.

\item The {\em external} part of this plan can then be posted which
  eventually (at the end of the current tick if these goals overlap
  the owner planning window) will be dispatched as goals triggering
  new deliberation on these reactors.

\end{enumerate}

@@

A core aspect of this reactor is that it takes advantage of the
specific sequencing enforced by the \rx architecture along with the
fact that only one $plan$ structure is maintained in the reactor and
shared between synchronization and planning to allow these two
processes to not only be interleaved but impact each other through the
way they manipulate this plan.

@@

Now, consider that at the next tick $\tau+1$ the reactor managing the
agent location was able to change the location state varaibale to the
\texttt{Going}. \rx informs our reactor of this new observation which
is added in the plan structure as a fact -- starting exactly at
$\tau+1$ and lastign for an unlknown duration which is at least 1
tick. As this new observation is integrated in the plan through
synchronization the solver can identify the similarity between this
new observation oand the next token in the existing plan. Which mean
that instead of making a blind exhaustive search of all the possible
implication the synchronization process starts its execution with the
assumption that these new observations are consitent with the current
plan maintained. As a result at the end of synchronization we will
have the plan depicted in Fig. \ref{fig:shop:exec1}. Which just
propagated the information provided by the new observation in the
exisiting plan (restricting in turn the start time of \texttt{Going}
and by propagtion of the constraints \texttt{Go} to be {\em exactly}
$\tau+1$). While focusing uniquely on the execution frontier ($\tau+1$
in this figure), the europa solver took benefit of the exisiting plan
to evaluate a solution that asumes that the observation is consistent
with what was decided inside this reactor.
